---
title: "R Notebook"
output: html_notebook
---

# Purpose

This script processes data are from experiments using arrays of artificial flowers conducted in the summer of 2022. Specifically, it cleans and reorganizes data from records of bees visiting the artificial arrays of flowers on six mornings (2 trials).


# Set up working directory

```{r, setup} 
library(knitr) 
opts_knit$set(root.dir = normalizePath('../'))
```


# Load libraries

```{r}
library(tidyverse)
```


# Step : Read in flower array files

```{r}
array_clove_0718 <- read.csv("data/0718_Clove_FeederVisits.csv", skip = 2)
array_peppermint_0718 <- read.csv("data/0718_Peppermint_FeederVisits.csv", skip = 2)
array_clove_0719 <- read.csv("data/0719_Clove_FeederVisits.csv", skip = 2)
array_peppermint_0719 <- read.csv("data/0719_Peppermint_FeederVisits.csv", skip = 2)
array_clove_0720 <- read.csv("data/0720_Clove_FeederVisits.csv", skip = 2)
array_peppermint_0720 <- read.csv("data/0720_Peppermint_FeederVisits.csv", skip = 2)

array_clove_0615 <- read.csv("data/0615_FeederVisits_Clove.csv")
array_peppermint_0615 <- read.csv("data/0615_FeederVisits_Peppermint.csv")
array_clove_0616 <- read.csv("data/0616_FeederVisits_Clove.csv")
array_peppermint_0616 <- read.csv("data/0616_FeederVisits_Peppermint.csv")
array_clove_0617 <- read.csv("data/0617_FeederVisits_Clove.csv")
array_peppermint_0617 <- read.csv("data/0617_FeederVisits_Peppermint.csv")
```

Make a list of these data frames to process more efficiently (the for loops that iterate over this list are based on code from this Stack Overflow answer https://stackoverflow.com/a/62216494):

```{r}
array_dfs <- list(array_clove_0718,
               array_peppermint_0718,
               array_clove_0719,
               array_peppermint_0719,
               array_clove_0720,
               array_peppermint_0720,
               array_clove_0615,
               array_peppermint_0615,
               array_clove_0616,
               array_peppermint_0616,
               array_clove_0617,
               array_peppermint_0617
               )
names(array_dfs) <- c("array_clove_0718",
               "array_peppermint_0718",
               "array_clove_0719",
               "array_peppermint_0719",
               "array_clove_0720",
               "array_peppermint_0720",
               "array_clove_0615",
               "array_peppermint_0615",
               "array_clove_0616",
               "array_peppermint_0616",
               "array_clove_0617",
               "array_peppermint_0617"
               )
```


# Step : Select only valid, double-marked bee columns

Function to select columns:

```{r}
# Remove columns that are not helpful for analyses
select_valid_bee_columns <- function(df){
  df <- df %>%
    select(!starts_with("Time")) %>%  # to keep hours:minutes into video
    select(!contains("Recruits")) %>%
    select(!contains("Notes")) %>%
    select(!ends_with("_"))  # to remove single-marked bees
  
  return(df)
}
```

Apply this to all data frames in the list

```{r}
for (i in 1:length(array_dfs)){
  array_dfs[[i]] <- select_valid_bee_columns(array_dfs[[i]])
  list2env(array_dfs,.GlobalEnv) # this is needed to modify the arrays globally rather than locally
}
```


# Step : Convert any text in bee observation columns to 1s

Some columns contain notes rather than 1s because I was making an educated guess about the true ID of the bee (we said an impossible ID in the tape). I think it is reasonable to assume that these guesses are mostly accurate.

```{r}
convert_text_to_ones <- function(df){
  
  df_colnames <- colnames(df)
  
  for (i in 2:length(df_colnames)){
  
      if (is_character(df[, i])){
    
        for (j in 1:length(df[, i])){
        
          if (df[j, i] == ""){
          
            df[j, i] <- NA
        
          } else {
          
            df[j, i] <- "1"  
        
          }
      }
      
      df[ , i] <- as.integer(df[ , i])
    
      }
  }

  return(df)
}
```

Apply this to the data frames in the list

```{r}
for (i in 1:length(array_dfs)){
  array_dfs[[i]] <- convert_text_to_ones(array_dfs[[i]])
  list2env(array_dfs,.GlobalEnv)
}
```


# Step : Convert flower array data frames so rows represent each bee sighting

Once we convert all of the data frames to a long format, we can combine them because they will have the same columns.

```{r}
convert_to_long_format <- . %>%
  pivot_longer(cols = !Vid_time_interval,
               names_to = "Bee",
               values_to = "Observed") %>%
  filter(Observed == 1)
```

Apply this to data frames

```{r}
for (i in 1:length(array_dfs)){
  array_dfs[[i]] <- convert_to_long_format(array_dfs[[i]])
  list2env(array_dfs,.GlobalEnv)
}
```


# Step : Add date, colony, scent, site, density, and flower number columns and combine all data frames into one

These variables are necessary for analyzing the data.

```{r}
array_clove_0718 <- array_clove_0718 %>%
  mutate(Date = "7/18/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0718 <- array_peppermint_0718 %>%
  mutate(Date = "7/18/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_clove_0719 <- array_clove_0719 %>%
  mutate(Date = "7/19/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0719 <- array_peppermint_0719 %>%
  mutate(Date = "7/19/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "High")
array_clove_0720 <- array_clove_0720 %>%
  mutate(Date = "7/20/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0720 <- array_peppermint_0720 %>%
  mutate(Date = "7/20/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "Low") %>%
  mutate(Flower_number = "Low")
array_clove_0615 <- array_clove_0615 %>%
  mutate(Date = "6/15/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0615 <- array_peppermint_0615 %>%
  mutate(Date = "6/15/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_clove_0616 <- array_clove_0616 %>%
  mutate(Date = "6/16/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "Low") %>%
  mutate(Flower_number = "Low")
array_peppermint_0616 <- array_peppermint_0616 %>%
  mutate(Date = "6/16/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_clove_0617 <- array_clove_0617 %>%
  mutate(Date = "6/17/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "High") 
array_peppermint_0617 <- array_peppermint_0617 %>%
  mutate(Date = "6/17/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")

combined_array_df <- rbind(array_clove_0615, array_peppermint_0615) %>%
  rbind(array_clove_0616) %>%
  rbind(array_peppermint_0616) %>%
  rbind(array_clove_0617) %>%
  rbind(array_peppermint_0617) %>%
  rbind(array_clove_0718) %>%
  rbind(array_peppermint_0718) %>%
  rbind(array_clove_0719) %>%
  rbind(array_peppermint_0719) %>%
  rbind(array_clove_0720) %>%
  rbind(array_peppermint_0720)
```


# Step : Add colony to Bee ID and remove WW.a_C (she was from a different colony)

```{r}
combined_array_df <- combined_array_df %>%
  mutate(Bee = paste(Bee, Colony, sep = "_")) %>%
  filter(Bee != "WW.a_C")
```


# Step : Modify data frame to add figure-relevant info

Adding day and constant/changing so they can be used in the figures:

```{r}
combined_array_df <- combined_array_df %>%
  # Add standardized "Day" variable for each trial
  mutate(Day = "0") %>% # initialize Day
  mutate(Day = ifelse(Date %in% c("6/15/22","7/18/22"), "1", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/16/22","7/19/22"), "2", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/17/22","7/20/22"), "3", Day)) %>%
  # Add Constant/Changing based on trial and scent info
  mutate(Treatment = "Null") %>% # initialize treatment
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Clove", "Changing", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Peppermint", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Clove", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Peppermint", "Changing", Treatment)) %>%
  # add trial number based on colony
  mutate(Trial = ifelse(Colony == "C", "Trial 1", "Trial 2")) %>%
  mutate(Setup = paste0("Density: ", Density, ", Number of flowers: ", Flower_number))
```


# Step : Fix Bee IDs that start with X

Some uniquely-marked bees in Trial one only had one mark on the abdomen, and we used ID such as "_B" to refer to them. Unfortunately, R automatically adds an X to the beginning of those column names so after we pivot to a long format we need to fix those IDs so that they will match the hive data.

```{r}
combined_array_df <- combined_array_df %>%
    mutate(Bee = gsub(pattern = "X",
                      replacement = "",
                      x = Bee))
```


# Step : Create "cleaned_data" folder if needed

```{r}
if (!dir.exists("cleaned_data")){
  dir.create("cleaned_data")
}
```


# Step : Add time into trial and write out full data frame

We need to be able to compare the time of the first observed visit to the flower array to the first observed trip back to the hive. At first I thought using the datetime class in lubridate would help, but it really seems simpler to just work with minutes into the trial (up to 2 hours in Trial 1 and up to 1 hour in Trial 2).

```{r}
combined_array_df <- combined_array_df %>%
  separate_wider_delim(cols = Vid_time_interval,
                       delim = ":",
                       names = c("Hour","Minute")) %>%
  mutate(Hour = as.double(Hour)) %>%
  mutate(Minute = as.double(Minute)) %>%
  mutate(Array_time = (Hour*60) + Minute)
```

And we'll save all of the visits so we can get cumulative bees visiting the arrays later.

```{r}
write.csv(combined_array_df,
          "cleaned_data/01_all_array_visits.csv",
          row.names = F)
```


# Step : Create new data frame with first sighting of each bee and length of time bee could have visited flower array

To compare with the hive data, what we really need is a summary with one row for each bee-day combination.

```{r}
combined_array_df <- combined_array_df %>%
  group_by(Colony, Trial, Date, Day, Scent, Treatment, Setup, Site, Density, Flower_number, Bee) %>%
  summarize(Array_time = min(Array_time)) %>%
  # then we can add the amount of time between when we first saw them and the end of the trial
  # which is important later for getting the dance rate
  mutate(Duration = ifelse(Colony == "7B", 60 - Array_time, 120 - Array_time))
```


# Step : Write out combined data frame to csv file

```{r}
write.csv(combined_array_df, 
          "cleaned_data/01_combined_flower_array_data.csv", 
          row.names = F)
```
