---
title: "R Notebook"
output: html_notebook
---

# Purpose

This script processes data are from experiments using arrays of artificial flowers conducted in the summer of 2022. Specifically, it cleans and reorganizes data from records of bees visiting the artificial arrays of flowers on six mornings (2 trials).


# Set up working directory

```{r, setup} 
library(knitr) 
opts_knit$set(root.dir = normalizePath('../'))
```


# Load libraries

```{r}
library(tidyverse)
```


# Read in flower array files

In both trials, the south array had 50 ul of clove bud essential oil added and the northeast array had 50 ul of peppermint oil essential oil added (hence the filenames).

```{r}
array_clove_0718 <- read.csv("data/0718_Clove_FeederVisits.csv", skip = 2)
array_peppermint_0718 <- read.csv("data/0718_Peppermint_FeederVisits.csv", skip = 2)
array_clove_0719 <- read.csv("data/0719_Clove_FeederVisits.csv", skip = 2)
array_peppermint_0719 <- read.csv("data/0719_Peppermint_FeederVisits.csv", skip = 2)
array_clove_0720 <- read.csv("data/0720_Clove_FeederVisits.csv", skip = 2)
array_peppermint_0720 <- read.csv("data/0720_Peppermint_FeederVisits.csv", skip = 2)

array_clove_0615 <- read.csv("data/0615_FeederVisits_Clove.csv")
array_peppermint_0615 <- read.csv("data/0615_FeederVisits_Peppermint.csv")
array_clove_0616 <- read.csv("data/0616_FeederVisits_Clove.csv")
array_peppermint_0616 <- read.csv("data/0616_FeederVisits_Peppermint.csv")
array_clove_0617 <- read.csv("data/0617_FeederVisits_Clove.csv")
array_peppermint_0617 <- read.csv("data/0617_FeederVisits_Peppermint.csv")
```

Make a list of these data frames to process more efficiently (the for loops that iterate over this list are based on code from this Stack Overflow answer https://stackoverflow.com/a/62216494):

```{r}
array_dfs <- list(array_clove_0718,
               array_peppermint_0718,
               array_clove_0719,
               array_peppermint_0719,
               array_clove_0720,
               array_peppermint_0720,
               array_clove_0615,
               array_peppermint_0615,
               array_clove_0616,
               array_peppermint_0616,
               array_clove_0617,
               array_peppermint_0617
               )
names(array_dfs) <- c("array_clove_0718",
               "array_peppermint_0718",
               "array_clove_0719",
               "array_peppermint_0719",
               "array_clove_0720",
               "array_peppermint_0720",
               "array_clove_0615",
               "array_peppermint_0615",
               "array_clove_0616",
               "array_peppermint_0616",
               "array_clove_0617",
               "array_peppermint_0617"
               )
```


# Select only valid, double-marked bee columns

Some columns in these data files counted unmarked bees or single-marked bees. Unfortunately, multiple bees had the same single mark so we could not analyze their behavior over time.

Function to select columns:

```{r}
# Remove columns that are not helpful for analyses
select_valid_bee_columns <- function(df){
  df <- df %>%
    select(!starts_with("Time")) %>%  # to keep hours:minutes into video
    select(!contains("Recruits")) %>%
    select(!contains("Notes")) %>%
    select(!ends_with("_")) %>%  # to remove single-marked bees
    select(!matches("^X$")) # to remove any extra blank columns
  
  return(df)
}
```

Apply this to all data frames in the list

```{r}
for (i in 1:length(array_dfs)){
  array_dfs[[i]] <- select_valid_bee_columns(array_dfs[[i]])
  list2env(array_dfs,.GlobalEnv) # this is needed to modify the arrays globally rather than locally
}
```


# Convert any text in bee observation columns to 1s

Some columns contain notes rather than 1s because I was making an educated guess about the true ID of the bee (we said an impossible ID in the recording). I think it is reasonable to assume that these guesses are accurate.

```{r}
convert_text_to_ones <- function(df){
  
  df_colnames <- colnames(df)
  
  for (i in 2:length(df_colnames)){
      
      # if the whole column contained only 1s and blanks, 
      # then it will have been read it in as numeric 
      # so we can skip that column
      if (is_character(df[, i])){
    
        for (j in 1:length(df[, i])){
        
          # if we didn't enter anything, 
          # then that bee was not IDed at the array in that minute
          if (df[j, i] == ""){
          
            df[j, i] <- NA
        
            # if we did enter something ("1" or text), then she was IDed
          } else {
          
            df[j, i] <- "1"  
        
          }
      }
      
      # And once we correct the blanks to NA and text to 1s, 
      # we need to convert the whole column to integer
      df[ , i] <- as.integer(df[ , i])
    
      }
  }

  return(df)
}
```

Apply this to the data frames in the list

```{r}
for (i in 1:length(array_dfs)){
  array_dfs[[i]] <- convert_text_to_ones(array_dfs[[i]])
  list2env(array_dfs,.GlobalEnv)
}
```


# Convert flower array data frames so rows represent each bee sighting

Once we convert all of the data frames to a long format, we can combine them because they will have the same columns.

```{r}
convert_to_long_format <- . %>%
  pivot_longer(cols = !Vid_time_interval,
               names_to = "Bee",
               values_to = "Observed") %>%
  filter(Observed == 1)  # we only need the times they were observed
```

Apply this to data frames

```{r}
for (i in 1:length(array_dfs)){
  array_dfs[[i]] <- convert_to_long_format(array_dfs[[i]])
  list2env(array_dfs,.GlobalEnv)
}
```


# Add date, colony, scent, site, density, and flower number columns and combine all data frames into one

These variables are necessary for analyzing the data later.

```{r}
# Trial 1:
  # South array
    # Clove scent
    # 7/18-> High density (full trays), low flower number (67)
    # 7/19-> Low density (1/9th full trays), low flower number (67)
    # 7/20-> High density (full trays), high flower number (603)
  # Northeast array
    # Peppermint scent
    # 7/18 to 7/20-> High density (full trays), low flower number (67) 
array_clove_0615 <- array_clove_0615 %>%
  mutate(Date = "6/15/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0615 <- array_peppermint_0615 %>%
  mutate(Date = "6/15/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_clove_0616 <- array_clove_0616 %>%
  mutate(Date = "6/16/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "Low") %>%
  mutate(Flower_number = "Low")
array_peppermint_0616 <- array_peppermint_0616 %>%
  mutate(Date = "6/16/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_clove_0617 <- array_clove_0617 %>%
  mutate(Date = "6/17/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "High") 
array_peppermint_0617 <- array_peppermint_0617 %>%
  mutate(Date = "6/17/22") %>%
  mutate(Colony = "C") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")

# Trial 2:
  # South array
    # Clove scent
    # 7/18 to 7/20-> High density (full trays), low flower number (67)  
  # Northeast array
    # Peppermint scent
    # 7/18-> High density (full trays), low flower number (67)
    # 7/19-> High density (full trays), high flower number (603)
    # 7/20-> Low density (1/9th full trays), low flower number (67)
array_clove_0718 <- array_clove_0718 %>%
  mutate(Date = "7/18/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0718 <- array_peppermint_0718 %>%
  mutate(Date = "7/18/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_clove_0719 <- array_clove_0719 %>%
  mutate(Date = "7/19/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0719 <- array_peppermint_0719 %>%
  mutate(Date = "7/19/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "High")
array_clove_0720 <- array_clove_0720 %>%
  mutate(Date = "7/20/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Clove") %>%
  mutate(Site = "South") %>%
  mutate(Density = "High") %>%
  mutate(Flower_number = "Low")
array_peppermint_0720 <- array_peppermint_0720 %>%
  mutate(Date = "7/20/22") %>%
  mutate(Colony = "7B") %>%
  mutate(Scent = "Peppermint") %>%
  mutate(Site = "Northeast") %>%
  mutate(Density = "Low") %>%
  mutate(Flower_number = "Low")

combined_array_df <- rbind(array_clove_0615, array_peppermint_0615) %>%
  rbind(array_clove_0616) %>%
  rbind(array_peppermint_0616) %>%
  rbind(array_clove_0617) %>%
  rbind(array_peppermint_0617) %>%
  rbind(array_clove_0718) %>%
  rbind(array_peppermint_0718) %>%
  rbind(array_clove_0719) %>%
  rbind(array_peppermint_0719) %>%
  rbind(array_clove_0720) %>%
  rbind(array_peppermint_0720)

```


# Add colony to Bee ID and remove WW.a_C (she was from a different colony)

Because some color combinations were used in both trials, we need to make all bee IDs unique by adding the colony to the ID.

```{r}
combined_array_df <- combined_array_df %>%
  mutate(Bee = paste(Bee, Colony, sep = "_")) %>%
  filter(Bee != "WW.a_C")
```


# Modify data frame to add figure-relevant info

Adding day and constant/changing variables so they can be used in the figures:

```{r}
combined_array_df <- combined_array_df %>%
  # Add standardized "Day" variable for each trial
  mutate(Day = "0") %>% # initialize Day
  mutate(Day = ifelse(Date %in% c("6/15/22","7/18/22"), "1", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/16/22","7/19/22"), "2", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/17/22","7/20/22"), "3", Day)) %>%
  # Add Constant/Changing based on trial and scent info
  mutate(Treatment = "Null") %>% # initialize treatment
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Clove", "Changing", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Peppermint", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Clove", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Peppermint", "Changing", Treatment)) %>%
  # add trial number based on colony
  mutate(Trial = ifelse(Colony == "C", "Trial 1", "Trial 2")) %>%
  mutate(Setup = paste0("Density: ", Density, ", Number: ", Flower_number))
```


# Fix Bee IDs that start with X

Some uniquely-marked bees in Trial one only had one mark on the abdomen, and we used IDs such as "_B" to refer to them. Unfortunately, R automatically adds an X to the beginning of those column names so after we pivot to a long format we need to fix those IDs so that they will match the hive data.

```{r}
combined_array_df <- combined_array_df %>%
    mutate(Bee = gsub(pattern = "X",
                      replacement = "",
                      x = Bee))
```


# Create "cleaned_data" folder if needed

For saving processed data files.

```{r}
if (!dir.exists("cleaned_data")){
  dir.create("cleaned_data")
}
```


# Add time into trial and write out full data frame

We need to be able to compare the time of the first observed visit to the flower array to the first observed trip back to the hive. At first I thought using the datetime class in lubridate would help, but it really seems simpler to just work with minutes into the trial (up to 120 minutes in Trial 1 and up to 60 minutes in Trial 2).

```{r}
combined_array_df <- combined_array_df %>%
  separate_wider_delim(cols = Vid_time_interval,
                       delim = ":",
                       names = c("Hour","Minute")) %>%
  mutate(Hour = as.double(Hour)) %>%
  mutate(Minute = as.double(Minute)) %>%
  mutate(Array_time = (Hour*60) + Minute)
```

And we'll save all of the visits so we can get cumulative bees visiting the arrays later.

```{r}
write.csv(combined_array_df,
          "cleaned_data/01_all_array_visits.csv",
          row.names = F)
```


# Create new data frame with first sighting of each bee and length of time bee could have visited flower array

To compare with the hive data, what we really need is a summary with one row for each bee-day combination.

```{r}
combined_array_df <- combined_array_df %>%
  group_by(Colony, Trial, Date, Day, Scent, Treatment, Setup, Site, Density, Flower_number, Bee) %>%
  summarize(Array_time = min(Array_time)) %>%
  # then we can add the amount of time between when we first saw them and the end of the trial
  # which is useful later if we want to calculate the dance rate
  mutate(Duration = ifelse(Colony == "7B", 60 - Array_time, 120 - Array_time))
```


# Write out combined data frame to csv file

```{r}
write.csv(combined_array_df, 
          "cleaned_data/01_combined_flower_array_data.csv", 
          row.names = F)
```
