---
title: "R Notebook"
output: html_notebook
---

# Set up working directory

```{r, setup} 
library(knitr) 
opts_knit$set(root.dir = normalizePath('../'))
```

# Load libraries

```{r}
library(tidyverse)
```

# Step : Read in hive csv files

```{r}
hive_0615 <- read.csv("data/0615_VideoAnalysis.csv")
hive_0616 <- read.csv("data/0616_VideoAnalysis.csv")
hive_0617 <- read.csv("data/0617_VideoAnalysis.csv")

hive_0718 <- read.csv("data/0718_VideoAnalysis.csv", skip = 1)
hive_0719 <- read.csv("data/0719_VideoAnalysis.csv", skip = 1)
hive_0720 <- read.csv("data/0720_VideoAnalysis.csv", skip = 1) %>%
  filter(Bee != "")  # for some reason there were many blank lines
```

Make a list of these data frames to process more efficiently (the for loops that iterate over this list are based on code from this Stack Overflow answer https://stackoverflow.com/a/62216494):

```{r}
hive_dfs <- list(hive_0615,
                      hive_0616,
                      hive_0617,
                      hive_0718,
                      hive_0719,
                      hive_0720
               )
names(hive_dfs) <- c("hive_0615",
                      "hive_0616",
                      "hive_0617",
                      "hive_0718",
                      "hive_0719",
                      "hive_0720"
               )
```

# Step : Select columns we will use

These include:

* Colony
* Date
* Bee
* Entrance minute
* Entrance second
* Danced
* Bouts columns (1-14)

```{r}
select_hive_columns <- . %>%
  select(Colony, 
         Date, 
         Bee, 
         Entrance_minute,
         Entrance_second,
         Danced,
         Bout_1,
         Bout_2,
         Bout_3,
         Bout_4,
         Bout_5,
         Bout_6,
         Bout_7,
         Bout_8,
         Bout_9,
         Bout_10,
         Bout_11,
         Bout_12,
         Bout_13,
         Bout_14)
```

```{r}
for (i in 1:length(hive_dfs)){
  hive_dfs[[i]] <- select_hive_columns(hive_dfs[[i]])
  list2env(hive_dfs,.GlobalEnv)
}
```


# Step : Combine hive data frames into one

```{r}
combined_hive_df <- rbind(hive_dfs[[1]], hive_dfs[[2]])
for (i in 3:length(hive_dfs)){
  combined_hive_df <- rbind(combined_hive_df, hive_dfs[[i]])
}
```


# Step : Convert bee IDs to standard format

Use periods instead of dashes to match the flower array ID format and add Colony to the bee ID to deal with repeated color combinations across trials of the experiment

```{r}
for (row in 1:length(combined_hive_df$Bee)){
    
  combined_hive_df$Bee[row] <- gsub(pattern = "-",
                        replacement = ".",
                        x = combined_hive_df$Bee[row])
}
  
combined_hive_df <- combined_hive_df %>%
  mutate(Bee = paste(Bee, Colony, sep = "_"))

```


# Step : Calculate decimal entrance time and total waggle runs

Then we need to calculate the decimal time of each trip entrance in the hive (and we might as well add the sum of the waggle runs as well):

```{r}
combined_hive_df <- combined_hive_df %>%
  mutate(Decimal_entrance_time = Entrance_minute + (Entrance_second/60)) %>%
  # for some reason, the dance bout columns were loaded in as logical vectors so we need to convert them to integers before summing them
  mutate(Bout_1 = as.integer(Bout_1)) %>% 
  mutate(Bout_2 = as.integer(Bout_2)) %>% 
  mutate(Bout_3 = as.integer(Bout_3)) %>% 
  mutate(Bout_4 = as.integer(Bout_4)) %>% 
  mutate(Bout_5 = as.integer(Bout_5)) %>% 
  mutate(Bout_6 = as.integer(Bout_6)) %>% 
  mutate(Bout_7 = as.integer(Bout_7)) %>% 
  mutate(Bout_8 = as.integer(Bout_8)) %>% 
  mutate(Bout_9 = as.integer(Bout_9)) %>% 
  mutate(Bout_10 = as.integer(Bout_10)) %>% 
  mutate(Bout_11 = as.integer(Bout_11)) %>% 
  mutate(Bout_12 = as.integer(Bout_12)) %>% 
  mutate(Bout_13 = as.integer(Bout_13)) %>% 
  mutate(Bout_14 = as.integer(Bout_14)) %>% 
  mutate(Total_runs = rowSums(across(
                          c(Bout_1,
                          Bout_2,
                          Bout_3,
                          Bout_4,
                          Bout_5,
                          Bout_6,
                          Bout_7,
                          Bout_8,
                          Bout_9,
                          Bout_10,
                          Bout_11,
                          Bout_12,
                          Bout_13,
                          Bout_14)
                          ),
                          na.rm = T))
```


# Step : Fix Bee IDs that differed in flower array records

```{r}
combined_hive_df <- combined_hive_df %>%
  # OP.a is called OF.a at the feeder and that really is more
  # consistent with her appearance
  mutate(Bee = ifelse(Bee == "OP.a_7B", "OF.a_7B", Bee)) %>%
  # Unfortunately, GY and GP seem to be indistinguishable
  filter(Bee != "GP_7B") %>%
  filter(Bee != "GY_7B") %>%
  # Also, BWW is difficult to distinguish from BW and may have visited
  # both arrays at different times
  filter(Bee != "BWW_7B") %>%
  # WR was seen with pollen on her legs after visiting the feeder
  # so I think she may have been visiting multiple sources
  filter(Bee != "WR_7B")
```


# Step : Read in the flower array summary csv files created by 01_clean_array_data.Rmd:

```{r}
array_df <- read.csv("cleaned_data/combined_flower_array_data.csv")
```


# Step : Combine hive data with flower array data

```{r}
combined_hive_df <- combined_hive_df %>%
  filter(!is.na(Danced)) %>%  # to get rid of visits we decided were invalid
  full_join(array_df)
```

Then we can check that the bees that visited the hive but weren't recorded at a flower array make sense:

```{r}
combined_hive_df %>%
  filter(is.na(Decimal_time)) %>%
  group_by(Date, Bee) %>%
  summarize()
```

And we can remove them:

```{r}
combined_hive_df <- combined_hive_df %>%
  filter(!is.na(Decimal_time))
```

Now we can check how many dances were too early:

```{r}
combined_hive_df %>%
  filter(Danced == 1) %>%
  filter(Decimal_entrance_time < Decimal_time) %>%
  select(Date, Bee, Decimal_entrance_time, Decimal_time)
```

Based on this, I think it makes sense to give some leeway of 5 minutes (approximately one trip) in trial 2 because it was hard to keep track of every flower array visit.

```{r}
combined_hive_df <- combined_hive_df %>%
  filter(is.na(Decimal_entrance_time) | (Decimal_entrance_time - Decimal_time > -5)) # either the bee was never recorded in the hive or she was recorded at a reasonable time
```


# Step : Modify data frame to add figure-relevant info

Adding day and constant/changing so they can be used in the figures:

```{r}
combined_hive_df <- combined_hive_df %>%
  # Add standardized "Day" variable for each trial
  mutate(Day = "0") %>% # initialize Day
  mutate(Day = ifelse(Date %in% c("6/15/22","7/18/22"), "1", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/16/22","7/19/22"), "2", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/17/22","7/20/22"), "3", Day)) %>%
  # Add Constant/Changing based on trial and scent info
  mutate(Treatment = "Null") %>% # initialize treatment
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Clove", "Changing", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Peppermint", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Clove", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Peppermint", "Changing", Treatment)) %>%
  # add trial number based on colony
  mutate(Trial = ifelse(Colony == "C", "Trial 1", "Trial 2")) %>%
  mutate(Setup = paste0("Density: ", Density, ", Number of flowers: ", Flower_number))
```


# Step : Write data frame of dances to a csv file

All recorded hive visits:

```{r}
combined_hive_df %>%
  filter(!is.na(Danced)) %>%
write.csv(.,
          "cleaned_data/all_hive_visits.csv",
          row.names = F)
```

All recorded hive visits that included at least one waggle run:

```{r}
combined_hive_df %>%
  filter(Danced == 1) %>%
  write.csv(., 
          "cleaned_data/all_dances.csv", 
          row.names = F)
```

# Step : Make a summary by bee and date

```{r}
hive_summary <- combined_hive_df %>%
  group_by(Colony, Trial, Date, Day, Scent, Treatment, Setup, Site, Density, Flower_number, Bee) %>%
  summarize(Dances = sum(Danced, na.rm = F), 
            Dance_rate = ifelse(sum(Danced, na.rm = F) > 0 & mean(Duration) < 0, 1, sum(Danced, na.rm = F) / mean(Duration/60)),
            Waggle_runs = sum(Total_runs, na.rm = F)) %>%
  mutate(Dances = ifelse(is.na(Dances), 0, Dances)) %>%
  mutate(Dance_rate = ifelse(is.na(Dance_rate), 0, Dance_rate)) %>%
  mutate(Waggle_runs = ifelse(is.na(Waggle_runs), 0, Waggle_runs))
```

Write the summary to a csv file:

```{r}
write.csv(hive_summary,
          "cleaned_data/per_bee_summary.csv",
          row.names = F)
```
