---
title: "R Notebook"
output: html_notebook
---

# Set up working directory

```{r, setup} 
library(knitr) 
opts_knit$set(root.dir = normalizePath('../'))
```


# Load libraries
```{r}
library("tidyverse")
library("lme4")
library("nlme")
library("DHARMa")
library("car")
```


# Step : Read in data files

```{r}
# This file has one row per dance
dances_df <- read.csv("cleaned_data/all_dances.csv")

# This file has one row per uniquely-IDed bee
bees_df <- read.csv("cleaned_data/per_bee_summary.csv")

```

# Step : Per bee dances model

```{r}
model_dances <- glmer.nb(Dances ~ Density + Flower_number + Colony + (1|Date),
                         data = bees_df)

summary(model_dances)
```

Testing model fit with DHARMa package:

```{r}
sim_model_dances <- simulateResiduals(model_dances, refit = T)
testDispersion(sim_model_dances)
testUniformity(sim_model_dances)
testOutliers(sim_model_dances)
```

I don't know how best to fix this because a Poisson regression was worse.


# Step : Per bee dance_rate model

```{r}
model_dance_rate <- lme(Dance_rate ~ Density + Flower_number + Colony,
                        random = ~1|Date,
                        data = bees_df)

summary(model_dance_rate)
```

Testing model fit with plot:

```{r}
qqnorm(resid(model_dance_rate))
qqline(resid(model_dance_rate))
```
The DHARMa package does not seem to work for lme models. This qqplot looks worrying, especially at the lower theoretical values. I tried transformations (sqrt(x), log(x + 1)), but they don't seem to help.

# Step : Per dance waggle run model

```{r}
model_waggle_runs <- glmer.nb(Total_runs ~ Density + Flower_number + Colony + (1|Bee/Date),
                              data = dances_df)

summary(model_waggle_runs)
```

Testing model fit with DHARMa package:

```{r}
sim_model_waggle_runs <- simulateResiduals(model_waggle_runs, refit = T)
testDispersion(sim_model_waggle_runs)
testUniformity(sim_model_waggle_runs)
testOutliers(sim_model_waggle_runs)
```

This seems reasonable to me. No significant over/underdispersion and the qqplot looks good. Excluding Colony does not appear to change the result but adding colony as a random effect (bee nested in date nested in colony) causes the model to not converge.

