---
title: "R Notebook"
output: html_notebook
---

# Set up working directory

```{r, setup} 
library(knitr) 
opts_knit$set(root.dir = normalizePath('../'))
```


# Load libraries
```{r}
library("tidyverse")
```


# Step : Read in data files

```{r}
# This file has one row per dance
dances_df <- read.csv("cleaned_data/all_dances.csv")

# This file has one row per uniquely-IDed bee
bees_df <- read.csv("cleaned_data/per_bee_summary.csv")

```

# Step : Modify data frames to add figure-relevant info

Adding day and constant/changing so they can be used in the figures:

```{r}
# Helper function
add_info <- . %>%
  # Add standardized "Day" variable for each trial
  mutate(Day = "0") %>% # initialize Day
  mutate(Day = ifelse(Date %in% c("6/15/22","7/18/22"), "1", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/16/22","7/19/22"), "2", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/17/22","7/20/22"), "3", Day)) %>%
  # Add Constant/Changing based on trial and scent info
  mutate(Treatment = "Null") %>% # initialize treatment
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Clove", "Changing", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Peppermint", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Clove", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Peppermint", "Changing", Treatment)) %>%
  # add trial number based on colony
  mutate(Trial = ifelse(Colony == "C", "Trial 1", "Trial 2"))
```

```{r}
bees_df <- add_info(bees_df)

dances_df <- add_info(dances_df)
```


# Step : Create "figures" folder if needed

```{r}
if (!dir.exists("figures")){
  dir.create("figures")
}
```


# Step : Make boxplot figures

Dances per bee

```{r}
bees_df %>%
  mutate(Setup = paste0("Density: ", Density, " , Number of flowers: ", Flower_number)) %>%
  ggplot(aes(x = Day, y = Dances, fill = Setup)) +
  geom_boxplot() +
  labs(y = "Dances performed per bee") +
  facet_grid(Trial~Treatment, scales = "free_y") +
  theme_classic(base_size = 16)

ggsave("figures/Boxplot_dances_per_bee.jpg", 
       device="jpeg",
       width = 10,
       height = 8,
       units = "in",
       dpi = 300)
```

Dances per bee per hour after she was first seen visiting an artificial flower array

```{r}
bees_df %>%
  mutate(Setup = paste0("Density: ", Density, " , Number of flowers: ", Flower_number)) %>%
  ggplot(aes(x = Day, y = Dance_rate, fill = Setup)) +
  geom_boxplot() +
  labs(y = "Dances performed per bee per hour") +
  facet_grid(Trial~Treatment, scales = "free_y") +
  theme_classic(base_size = 16)

ggsave("figures/Boxplot_dance_rate_per_bee.jpg", 
       device="jpeg",
       width = 10,
       height = 8,
       units = "in",
       dpi = 300)
```

Waggle runs per dance

```{r}
dances_df %>%
  mutate(Setup = paste0("Density: ", Density, " , Number of flowers: ", Flower_number)) %>%
  ggplot(aes(x = Day, y = Total_runs, fill = Setup)) +
  geom_boxplot() +
  labs(y = "Waggle runs per dance") +
  facet_grid(Trial~Treatment, scales = "free_y")+
  theme_classic(base_size = 16)

ggsave("figures/Boxplot_waggle_runs_per_dance.jpg", 
       device="jpeg",
       width = 10,
       height = 8,
       units = "in",
       dpi = 300)
```