---
title: "R Notebook"
output: html_notebook
---

# Set up working directory

```{r, setup} 
library(knitr) 
opts_knit$set(root.dir = normalizePath('../'))
```


# Load libraries
```{r}
library("tidyverse")
```


# Step : Read in data files

```{r}
# This file has one row per hive visit
dances_df <- read.csv("cleaned_data/all_dances.csv")

```


# Step : Modify data frames to add figure-relevant info

Adding day and constant/changing so they can be used in the figures:

```{r}
# Helper function
add_info <- . %>%
  # Add standardized "Day" variable for each trial
  mutate(Day = "0") %>% # initialize Day
  mutate(Day = ifelse(Date %in% c("6/15/22","7/18/22"), "1", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/16/22","7/19/22"), "2", Day)) %>%
  mutate(Day = ifelse(Date %in% c("6/17/22","7/20/22"), "3", Day)) %>%
  # Add Constant/Changing based on trial and scent info
  mutate(Treatment = "Null") %>% # initialize treatment
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Clove", "Changing", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "C" & Scent == "Peppermint", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Clove", "Constant", Treatment)) %>%
  mutate(Treatment = ifelse(Colony == "7B" & Scent == "Peppermint", "Changing", Treatment)) %>%
  # add trial number based on colony
  mutate(Trial = ifelse(Colony == "C", "Trial 1", "Trial 2"))
```

```{r}
dances_df <- add_info(dances_df) %>%
  mutate(Integer_time = floor(Decimal_entrance_time)) %>%
  mutate(Cumul_bees = 0)  # so we can fill in this column below
```


# Step : Calculate cumulative bees, dances, and waggle runs

```{r}
# Adding cumulative bees (I think!) requires a set of nested for loops

cumulative_df <- data.frame(Trial = character(),
                 Day = character(),
                 Treatment = character(),
                 Density = character(),
                 Flower_number = character(),
                 Minute = integer(),
                 Cumul_bees = integer(),
                 Cumul_dances = integer(),
                 Cumul_waggle_runs = integer())

## We need to create one list of bees for every trial, day, treatment combination
## We use these lists to create a smaller data frame with cumulative bees added
## and then we need to add each of these temporary data frames to a larger
## data frame
trials <- unique(dances_df$Trial)
days <- unique(dances_df$Day)
treatments <- unique(dances_df$Treatment)

for (trial_i in trials){
  if (trial_i == "Trial 1"){
    minutes <- 1:119
  } else {
    minutes <- 1:59
  }
  for (day_i in days){
    for (treatment_i in treatments){
      # get that subset of rows
      temp_df <- dances_df %>%
        filter(Trial == trial_i & Day == day_i & Treatment == treatment_i)
      curr_density <- temp_df$Density[1]
      curr_flower_number <- temp_df$Flower_number[1]
      # create a temporary df to store the cumulative values
      temp_cumul_df <- data.frame(Trial = rep(x = trial_i, times = length(minutes)),
                                  Day = rep(x = day_i, times = length(minutes)),
                                  Treatment = rep(x = treatment_i, times = length(minutes)),
                                  Density = rep(x = curr_density, times = length(minutes)),
                                  Flower_number = rep(x = curr_flower_number, times = length(minutes)),
                                  Minute = minutes,
                                  Cumul_bees = rep(x = 0, times = length(minutes)),
                                  Cumul_dances = rep(x = 0, times = length(minutes)),
                                  Cumul_waggle_runs = rep(x = 0, times = length(minutes)))
      for (row in 1:length(minutes)){
        # save the minute for the current row
        curr_minute <- row
        # get a subset of temp_df
        temp_df_row <- temp_df %>%
          filter(Integer_time <= curr_minute)
        # get a vector of all bees that entered during that minute
        curr_bees_list <- unique(temp_df_row$Bee)
        # add the cumulative totals to the temporary data frame
        temp_cumul_df$Cumul_bees[row] <- length(curr_bees_list)
        temp_cumul_df$Cumul_dances[row] <- sum(temp_df_row$Danced)
        temp_cumul_df$Cumul_waggle_runs[row] <- sum(temp_df_row$Total_runs)
      }
      # then add the temporary df to the df we initialized above
      cumulative_df <- rbind(cumulative_df, temp_cumul_df)
    }
  }
}
```


# Step : Create "figures" folder if needed

```{r}
if (!dir.exists("figures")){
  dir.create("figures")
}
```


# Step : Making cumulative figures

Unique bees who danced

```{r}
cumulative_df %>%
  mutate(Setup = paste0("Density: ", Density, " , Number of flowers: ", Flower_number)) %>%
  ggplot(aes(x = Minute, y = Cumul_bees, color = Setup, linetype = Day)) +
  scale_linetype_manual(values = c("twodash", "solid", "dotted")) +
  geom_line(linewidth = 1) +
  labs(y = "Cumulative bees that danced") +
  facet_grid(Treatment~Trial, scales = "free_x") +
  theme_light(base_size = 16)

ggsave("figures/Lineplot_cumulative_dancing_bees.jpg", 
       device="jpeg",
       width = 12,
       height = 8,
       units = "in",
       dpi = 300)
  
```

Dances performed

```{r}
cumulative_df %>%
  mutate(Setup = paste0("Density: ", Density, " , Number of flowers: ", Flower_number)) %>%
  ggplot(aes(x = Minute, y = Cumul_dances, color = Setup, linetype = Day)) +
  scale_linetype_manual(values = c("twodash", "solid", "dotted")) +
  geom_line(linewidth = 1) +
  labs(y = "Cumulative dances") +
  facet_grid(Treatment~Trial, scales = "free_x") +
  theme_light(base_size = 16)


ggsave("figures/Lineplot_cumulative_dances.jpg", 
       device="jpeg",
       width = 12,
       height = 8,
       units = "in",
       dpi = 300)
```

Waggle runs performed

```{r}
cumulative_df %>%
  mutate(Setup = paste0("Density: ", Density, " , Number of flowers: ", Flower_number)) %>%
  ggplot(aes(x = Minute, y = Cumul_waggle_runs, color = Setup, linetype = Day)) +
  scale_linetype_manual(values = c("twodash", "solid", "dotted")) +
  geom_line(linewidth = 1) +
  labs(y = "Cumulative waggle runs") +
  facet_grid(Treatment~Trial, scales = "free_x") +
  theme_light(base_size = 16)
  

ggsave("figures/Lineplot_cumulative_waggle_runs.jpg", 
       device="jpeg",
       width = 12,
       height = 8,
       units = "in",
       dpi = 300)
```

